---
title: "00 R, Rmarkdown, code, and {tidyverse}: <br> A whirlwind tour"
author: 
  - "STAT 406"
  - "Daniel J. McDonald"
date: 'Last modified - `r Sys.Date()`'
---

```{r setup, include=FALSE}
source("rmd_config.R")
```

class: middle, inverse, center

# The basics

---
## Tour of Rstudio

Things to note

1. Console
2. Terminal
3. Scripts, .Rmd, Knit
4. Files, Projects
5. Getting help
6. Environment, Git

---
## Simple stuff

.pull-left[
### Vectors:
```{r}
x <- c(1, 3, 4)
x[1]
x[-1]
rev(x)
c(x, x)
```
]

.pull-right[
### Matrices:
```{r}
x <- matrix(1:25, nrow = 5, ncol = 5)
x[1,]
x[,-1]
x[c(1,3),  2:3]
```

]

---
## Simple stuff

.pull-left[
### Lists
```{r}
(l <- list(a = letters[1:2], b = 1:4, c = list(a = 1)))
l$a
l$c$a
l["b"]
```
]

.pull-right[
### Data frames
```{r}
(dat <- data.frame(z = 1:5, b = 6:10, c = letters[1:5]))
class(dat)
dat$b
dat[1,]
```

.secondary[Data frames are sort-of lists and sort-of matrices]
]

---

## Tibbles

.secondary[These are `{tidyverse}` data frames]

```{r}
(dat2 <- tibble(z = 1:5, b = z + 5, c = letters[z]))
class(dat2)
```

We'll return to classes in a moment. A `tbl_df` is a "subclass" of `data.frame`.

Anything that `data.frame` can do, `tbl_df` can do (better).

For instance, the printing is more informative.

Also, you can construct one by referencing previously constructed columns.
---
class: middle, inverse, center

# Functions

---

## Understanding signatures

```{r echo=FALSE}
sig <- sig::sig
```

```{r}
sig(lm)
sig(`+`)
sig(ggplot2::layer)
sig(epipredict::layer)
sig(rnorm)
```

---

## These are all the same

```{r}
set.seed(12345)
rnorm(3)
set.seed(12345)
rnorm(n = 3, mean = 0)
set.seed(12345)
rnorm(3, 0, 1)
set.seed(12345)
rnorm(sd = 1, n = 3, mean = 0)
```

* Functions can have default values.
* You may, but don't have to, name the arguments
* If you name them, you can pass them out of order (but you shouldn't).

---

## Write lots of functions. I can't emphasize this enough.

.pull-left[
```{r functions}
f <- function(arg1, arg2, arg3 = 12, ...) {
  stuff <- arg1 * arg3
  stuff2 <- stuff + arg2
  plot(arg1, stuff2, ...)
  return(stuff2)
}
x <- rnorm(100)
```
]

.pull-right[
```{r plot-it}
y1 <- f(x, 3, 15, col = 4, pch = 19)
```
]

---

## Outputs vs. Side effects

.pull-left[
* Side effects are things a function does, outputs can be assigned to variables
* A good example is the `hist` function
* You have probably only seen the side effect which is to plot the histogram
```{r,fig.align='center', fig.height=5}
my_histogram <- hist(rnorm(1000))
```
]

.pull-right[
```{r}
my_histogram
class(my_histogram)
```

]

---

### When writing functions, program defensively, ensure behaviour

.pull-left[
```{r, error=TRUE}
incrementer <- function(x, inc_by = 1) return(x + 1)
incrementer(2)
incrementer(1:4)
incrementer("a")
```

--

```{r, error=TRUE}
incrementer <- function(x, inc_by = 1) {
  stopifnot(is.numeric(x))
  return(x + 1)
}
incrementer("a")
```

]

--

.pull-right[
```{r, error=TRUE}
incrementer <- function(x, inc_by = 1) {
  if (!is.numeric(x)) stop("`x` should be numeric")
  return(x + 1)
}
incrementer("a")
incrementer(2, -3) ## oops!
incrementer <- function(x, inc_by = 1) {
  if (!is.numeric(x)) stop("`x` should be numeric")
  return(x + inc_by)
}
incrementer(2, -3)
```
]

---

## How to keep track

.pull-left[
```{r}
library(testthat)
incrementer <- function(x, inc_by = 1) {
  if (!is.numeric(x)) stop("`x` should be numeric")
  if (!is.numeric(inc_by)) 
    stop("`inc_by` should be numeric")
  return(x + inc_by)
}
expect_error(incrementer("a"))
expect_equal(incrementer(1:3), 2:4)
expect_equal(incrementer(2, -3), -1)
expect_error(incrementer(1, "b"))
```

```{r, error=TRUE}
expect_identical(incrementer(1:3), 2:4)
```
]

--

.pull-right[
```{r}
is.integer(2:4)
is.integer(incrementer(1:3))
expect_identical(incrementer(1:3, 1L), 2:4)
```
]

--

.emphasis[
If you copy something, write a function.

Validate your arguments.

To ensure proper functionality, write tests to check if inputs result in predicted outputs.
]


---
class: middle, inverse, center

# Classes and methods

---

## Classes

.pull-left[
We saw some of these earlier:

```{r}
tib <- tibble(
  x1 = rnorm(100), 
  x2 = rnorm(100), 
  y = x1 + 2 * x2
)
mdl <- lm(y ~ ., data = tib )
class(tib)
class(mdl)
```

The class allows for the use of "methods"

```{r}
print(mdl)
```
]

--

.pull-right[
`R` "knows what to do" when you `print()` an object of class `"lm"`.

`print()` is called a "generic" function. You can create "methods" that get dispatched.

For any generic, `R` looks for a method for the class.

If available, it calls that function.

```{r}

sloop::s3_dispatch(print(incrementer))
sloop::s3_dispatch(print(tib))
sloop::s3_dispatch(print(mdl))
```
]

---


## R-Geeky But Important

There are .secondary[lots] of generic functions in `R`

Common ones are `print()`, `summary()`, and `plot()`.

Also, lots of important statistical modelling concepts:
`residuals()` `coefficients()` 

(In `python`, these work the opposite way: `obj.residuals`. The dot after the object accesses methods defined for that type of object. But the dispatch behaviour is less robust.)  

* The convention is
that the specialized function is named `method.class`, e.g., `summary.lm`.

* If no specialized function is defined, R will try to use `method.default`.

For this reason, `R` programmers try to avoid `.` in names of functions or objects.

--

### Wherefore methods?


* The advantage is that you don't have to learn a totally
new syntax to grab residuals or plot things

* You just use `residuals(mdl)` whether `mdl` comes from `lm`
could have been done two centuries ago, or a Batrachian Emphasis Machine
which won't be invented for another five years.  

* The one draw-back is the help pages for the generic methods tend
to be pretty vague

* Compare `?summary` with `?summary.lm`.  


---
class: middle, inverse, center

# Environments 

---

## Different environments

* These are often tricky, but are very common.

* Most programming languages have this concept in one way or another.

* In `R` code run in the Console produces objects in the "Global environment"

* You can see what you create in the "Environment" tab.

* But there's lots of other stuff.

* Many packages are automatically loaded at startup, so you have access to the functions and data inside

For example `mean()`, `lm()`, `plot()`, `iris` (technically `iris` is lazy-loaded, meaning it's not in memory until you call it, but it is available)

--

<hr>

* Other packages require you to load them with `library(pkg)` before their functions are available.

* But, you can call those functions by prefixing the package name `ggplot2::ggplot()`.

* You can also access functions that the package developer didn't "export" for use with `:::` like `dplyr:::as_across_fn_call()`

---

## Other issues with environments

.pull-left[
As one might expect, functions create an environment inside the function.
```{r}
z <- 1
fun <- function(x) {
  z <- x
  print(z)
  invisible(z)
}
fun(14)
```

More non-trivial cases are `data-masking` environments.

```{r, error=TRUE}
tib <- tibble(
  x1 = rnorm(100), 
  x2 = rnorm(100), 
  y = x1 + 2 * x2
)
mdl <- lm(y ~ x2, data = tib)
x2
```
* `lm()` looks "inside" the `tib` to find `y` and `x2`
* The data variables are added to the `lm` environment
]

--

.pull-right[

.emphasis[
When Knit, `.Rmd` files run in their OWN environment.
]

They are run from top to bottom, with code chunks depending on previous

This makes them reproducible.

Jupyter notebooks don't do this.

Objects in your local environment are not available in the `.Rmd`

Objects in the `.Rmd` are not available locally.

.emphasis[
The most frequent error I see is:

* running chunks individually, 1-by-1, and it works
* Knitting, and it fails

The reason is almost always that the chunks refer to objects in the Environment that don't exist in the `.Rmd`
]
]

---
class: middle

.alert[
### This error also happens because:

* `library()` calls were made globally but not in the `.Rmd` so the packages aren't loaded

* paths to data or other objects are not relative to the `.Rmd` in your file system (they must be)

]


---
class: middle, inverse, center

# Debugging

---

## How to fix code

* If you're using a function in a package, start with `?function` to see the help
    * Make sure you're calling the function correctly.
    * Try running the examples.
    * paste the error into Google (if you share the error on Slack, I often do this first)
    * Go to the package website if it exists, and browse around
    
* If your `.Rmd` won't Knit
    * Did you make the mistake on the last slide?
    * Did it Knit before? Then the bug is in whatever you added.
    * Did you never Knit it? Why not?
    * Call `rstudioapi::restartSession()`, then run the Chunks 1-by-1
    
* Adding `browser()`
    * Only useful with your own functions.
    * Open the script with the function, and add `browser()` to the code somewhere
    * Then call your function.
    * The execution will Stop where you added `browser() and you'll have access to the local objects to play around

---

## Reproducible examples

.emphasis[
Question I get on Slack that I hate:

"I ran the code like you had on Slide 39, but it didn't work."
]

If you want to ask me why the code doesn't work, you need to show me what's wrong.

.emphasis[
**Don't just paste a screenshot!** 

Unless you get lucky, I won't be able to figure it out from that. And we'll both get frustrated.
]

What you need is a Reproducible Example or `reprex`.

This is a small chunk of code that runs in it's own environment and produces the error.

The best way to do this is with the `{reprex}` package.

---
## Reproducible examples

### How it works

1. Open a new `.R` script.

1. Paste your buggy code in the file (no need to save)

1. Edit your code to make sure it's "enough to produce the error" and nothing more.

1. Copy your code.

1. Call `reprex::reprex(venue = "r")` from the console. This will run your code in a new environment and show the result in the Viewer tab. Does it create the error you expect?

1. If it creates other errors, that may be the problem. You may fix the bug on your own!

1. The Output is now on your clipboard. Go to Slack and paste it in a message. Then press `Cmd+Shift+Enter` (on Mac) or `Ctrl+Shift+Enter` (Windows/Linux). Under Type, select `R`.

1. Send the message, perhaps with more description and an SOS emoji.

---
class: middle, inverse, center

# Understanding `{tidyverse}`